Multi-level modeling with Openflexo/\FML

### 2021-09-22 responseLetter.tex

% Title:    A LaTeX Template For Responses To a Referees' Reports
% Author:   Petr Zemek <s3rvac@gmail.com>
% Homepage: https://blog.petrzemek.net/2016/07/17/latex-template-for-responses-to-referees-reports/
% License:  CC BY 4.0 (https://creativecommons.org/licenses/by/4.0/)
\documentclass[10pt]{article}

% Allow Unicode input (alternatively, you can use XeLaTeX or LuaLaTeX)
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{todonotes}

\usepackage{microtype,xparse,tcolorbox}
\newenvironment{reviewer-commentX}{}{}
\tcbuselibrary{skins}
\tcolorboxenvironment{reviewer-comment }{empty,
  left = 1em, top = 1ex, bottom = 1ex,
  borderline west = {2pt} {0pt} {black!20},
}
\ExplSyntaxOn
\NewDocumentEnvironment {response} { +m O{black!20} } {
  \IfValueT {#1} {
    \begin{reviewer-commentX}
      \setlength\parindent{2em}
      \noindent
      \ttfamily #1
    \end{reviewer-commentX}
  }
  \par\noindent\ignorespaces\color{blue}
} { \bigskip\par }

@@

\begin{response}{In Figure 5, the base metamodel for representing processes is shown - however it is not linked to the concepts you introduced before. Would it be possible to maybe graphically highlight how it relates to the metamodel in Figure 2? Besides, as multi-level modeling is much different to traditional modeling, you may want to consider using a different notation rather than UML class diagrams - as these are commonly known for traditional modeling approaches. When reading the paper, I actually marked the references to types in yellow in my copy - ie from ProcessType to TaskType and from ActorType to ActorType (on the attribute levels) - this helped me to see more clearly where you have references from attributes to types - but that's just my way of doing it.} 
TODO.
\end{response}

\begin{response}{Regarding Figure 6 and Figure 7: In Figure 7 you show the process incl. sequence connectors - however in Figure 6 these seem to be missing - maybe to reduce complexity? I would suggest to probably focus on a subset of the process from Figure 7 and also show how the connectors are realized. To me this would increase the comprehensibility / consistency. Ie I could then infer exactly, how elements from the concrete model have been realized in the background.} 
TODO.
\end{response}

\begin{response}{in Figure 9 you show the model editor. Could you please also add a screenshot of the metamodel editor?} 
TODO.
\end{response}

\begin{response}{There are some minor typos throughout the manuscript (eg also regarding the reference Jeusfeld2019; choose -> chose; axis -> axes; aN FML virtual model...; etc.)} 
TODO.
choose $\rightarrow$ chose OK ; axis $\rightarrow$  axes OK ; je ne comprends pas pour Jeusfeld2019
(deeptelos2019) ni pour aN FML (je dirais ``a FML'', mais je puis me tromper)
\end{response}

@@

\pagebreak

\Reviewer{\#3}

\begin{response}{(1) The Type-Instance pattern is key to the authors approach to the challenge. However, the benefits of this pattern are not explored in the context of MLM. For example, the authors do not really explain approaches such as Clabjects or potency levels as they compare to the Type-Instance pattern (or the challenge). Why it the pattern better? Is it just a different approach? What can it do that other approaches cannot? Perhaps the benefit is that the Type-Instance pattern can be implemented in most modelling and implementation technologies whereas other approaches need special facilities? Finally, in terms of the pattern, why should anyone care - why would should they use the pattern?} 
TODO.
\end{response}

\begin{response}{(2) In section 1 the authors list 2 key reasons why a strict approach to modelling is sub-optimal. However, they do not establish the case that their approach uses these 2 features to any concrete benefit. It would be appropriate (perhaps at the end of the article) to revisit these features and to clearly demonstrate that their approach supports them and that they have concrete utility in some way in the context of the challenge.}

We have added a paragraph in the Conclusions section indicating how Openflexo/FML permitted us to overcome the mentioned limitations of the strict modeling paradigm and how this was used in order to build our solution to the Multi-Level Process challenge. Notably, we mention how the type/object pattern allowed us to use model element instances as types in order to achieve ontological instantiation.
\end{response}

### 2021-09-22 10-conclusions.tex
small change to the conclusions

We fulfill all the challenge requirements and we propose a tooling that demonstrates the usability of our solution.  We developed an ad hoc solution, including models and metamodels, thanks to the flexibility provided by the metametamodeling infrastructure offered by Openflexo.

>>>>

We fulfill all the challenge requirements and we propose a tooling that
demonstrates the usability of our solution. We developed an ad hoc solution,
including models and metamodels, thanks to the flexibility provided by the
metametamodeling infrastructure offered by Openflexo. \noteSM{This infrastructure helped us to overcome the limitations of the strict modeling paradigm mentioned in Section \ref{sec:introduction}, this is, the lack of support for different forms of classifications and for the duality type/object. In that sense, our solution seamlessly integrates the type-object pattern which allows us to dynamically create different \emph{instances} of model elements (e.g., \emph{TaskType}, among others) and use them for typing other model elements. This, together with the FML core support for (level agnostic) classical linguistic instantiation permitted us to merge linguistic and ontological instantiation in order to provide a satisfactory solution to the Multi-Level Process challenge.}

### 2021-09-22 3-introduction.tex
removing trailing whitespace
small change to the conclusions

>>>>

Our modeling approach is based on two levels. The meta level (\FML) provides the (linguistic) concepts that are used to define the model level (ontological). The ontological level is used to define the models and meta-models that are needed. \FML allows to define models as first class entities (which we call VirtualModel) and to link them together (by extension). In the \mpc case study, we exploit this possibility to:
\begin{itemize}
    \item Extend and adapt existing models, keeping the previous ones, to meet changing modeling needs,
    \item Align this hierarchy of models with the process of their evolution.
\end{itemize}
This methodological choice allows us to adapt the models and keep track of their evolution each time the needs evolve.

Contrary to a classical multi-level approach, in which the notion of multi-level is integrated in the meta-model and must therefore be respected, our approach allows to build multi-levels on demand, and with the appropriate form. As a limitation, without multilevel-specific operators/concepts, the "custom" construction must be done explicitly during the model development and evolution process.

Our answer to the challenge is therefore not an answer using a classical multilevel approach that would allow an efficiency comparison with another classical multilevel answer, but an approach that proposes another organization of multilevels through the translation of needs into levels over time and the possibility to build an evolving hierarchy of adapted models. This is made possible by (1) the reification of the notion of model and the possibility to interconnect them (specialization) and (2) the freedom of meta-modeling (low model conformity, flexible meta-modeling).

### 2021-09-22 5-analysis.tex
removing trailing whitespace
typos
small change to the conclusions

Our analysis of the use case leads to identify two main modeling axis (as presented in the figure \ref{fig:LinguisticAndOntologicInstantiation}):
\begin{itemize}
    \item The horizontal axis is characterized as the ontological instantiation axis, in the sense that the domain type definition (\ie \texttt{ProcessType}) is referenced by an instance definition (\ie \texttt{Process}). In our base metamodel, each domain type definition is referenced by its instance definition, as developed in the section \ref{sec:ProcessEnactment}

@@

\begin{figure}
    \centering
    \includegraphics[width=1.0 \columnwidth]{Figures/Instantiation.pdf}
    \caption{Linguistic and ontologic instantiation}
    \label{fig:LinguisticAndOntologicInstantiation}
\end{figure}

>>>>

Our analysis of the use case leads to identify two main modeling axes (as presented in the figure \ref{fig:LinguisticAndOntologicInstantiation}):
\begin{itemize}
    \item The horizontal axis is characterized as the ontological instantiation axis, in the sense that the domain type definition (\ie \texttt{ProcessType}) is referenced by an instance definition (\ie \texttt{Process}). In our base metamodel, each domain type definition is referenced by its instance definition, as developed in the section \ref{sec:ProcessEnactment}

@@

\begin{figure}
    \centering
    \includegraphics[width=1.0 \columnwidth]{Figures/Instantiation.pdf}
    \caption{Problem  Domain Linguistic and ontologic instantiation}
    \label{fig:LinguisticAndOntologicInstantiation}
\end{figure}
